\Title

### 2021-08-24 latex/paper/main.tex
Started writing the segment on achieving rotations in the proof.

We have proved that any acrobot constrained by 
our VNHC \eqref{eqn:acrobot-constraint} gains/loses energy on \(\mathcal{O}_1\).
We must now confirm that the acrobot will start rotating:
the definition of energy gain states that all orbits initialized in
\(\mathcal{O}_1\) will escape compact subsets of \(\mathcal{O}_1\) in finite
time, which is not enough to conclude that these orbits will also escape the
boundary of \(\mathcal{O}_1\) in finite time.

Let \(E_\pi\) be the level set with energy \(E(\pi,0)\), 
which forms the boundary of \(\mathcal{O}_1\). 
Define \(\bar{\mathcal{O}}_1 = \mathcal{O}_1 \cup E_\pi\) to be the closure of
\(\mathcal{O}_1\).
An orbit of the acrobot will only begin rotating if it escapes
\(\bar{\mathcal{O}}_1\) by crossing through \(E_\pi\), and eventually remains
outside this set.

Taking the Jacobian of the constrained dynamics
~\eqref{eqn:acrobot-constrained-dynamics} at the upright equilibrium 
\((q_u,p_u) = (\pi,0)\) yields
\[
    J = \begin{bmatrix}
        -\frac{6mgl\bar{q}_aI}{5} & \frac{1 - 2m^2gl^3\bar{q}_a^2 I^2}{5ml^2} \\
        3mgl & mgl\bar{q}_aI
    \end{bmatrix}
    ,
\]
which has characteristic polynomial
\[
    \det\left(\lambda \Id{2} - J\right)
    = \lambda^2 + \frac{mgl\bar{q}_a I}{5} \lambda - 3g
    .
\]
By Descartes' rule of signs, this polynomial always has exactly one root with
positive real part. 
The equilibrium \((\pi,0)\) is therefore unstable, so the stable manifold
\(\Pi^+\) of initial conditions converging to \((\pi,0)\) is one-dimensional,
and hence is of measure zero in \(\SxR\).

Using \(x(t) := (q_u(t),p_u(t))\) as shorthand, let
\(x(0) \in \mathcal{O}_1\) be a nonzero initial condition of the acrobot.
Suppose by way of contradiction that the orbit \(x(\R)\) is confined within 
\(\bar{\mathcal{O}}_1\), and does not exit through \(E_\pi\) into the rotation
zone.
Since \(\bar{\mathcal{O}}_1\) is compact, the Birkhoff Theorem \cite{birkhoff}
implies:
\begin{itemize}
    \item The positive limit set \(L_+\) of \(x(t)\) is non-empty, compact, and
        invariant.
    \item The solution \(x(t)\) asymptotically tends to \(L_+\).
\end{itemize}
Since the acrobot gains energy on \(\mathcal{O}_1\), the positive limit set of
\(x(t)\) must be the largest invariant subset of \(E_\pi\). 
Taking the limit as \(r \to \pi\) of
\eqref{eqn:theta-dot-nom-osc}--\eqref{eqn:theta-dot-limit},
we get \(\dot{\theta} \geq 0\) on \(E_\pi\) with equality if and only if
\(\theta \in \{0,\pi\}\).
Thus, the largest invariant subset of \(E_\pi\) is either the upright
equilibrium \(\{(\pi,0)\}\) or \(E_\pi\) itself.

Taking the derivative of \(E(q_u,p_u)\) at the \(p_u\)-axis yields
\(\dot{E}(0,p_u) = g/(5l) \sin(q_a)p_u\).
This is non-zero on \(E_\pi\), so \(E_\pi\) is not invariant and the positive
limit set of \(x(t)\) must be the set \(\{(\pi,0)\}\). 
Hence, \(x(t)\) converges to \((\pi,0)\), so \(x(0) \in \Pi^+\).
Since \(\Pi^+\) is a set of measure zero in \(\SxR\), almost every orbit
initialized in \(\mathcal{O}_1\) must escape at least once through \(E_\pi\)
into the rotation domain.
Moreover, almost every orbit must eventually remain outside
\(\bar{\mathcal{O}}_1\).
If not, the positive limit set would once again be
\(E_\pi\); by the same argument as above, that orbit would be in \(\Pi^+\).

We conclude that almost all orbits beginning in \(\mathcal{O}_1\) will, in
finite time, escape the closure of \(\mathcal{O}_1\) and remain in the rotation
domain forever after.

>>>>

We have proved that any acrobot constrained by 
our VNHC \eqref{eqn:acrobot-constraint} gains/loses energy on \(\mathcal{O}_1\).
We must now confirm that the acrobot will begin rotating, 
i.e., that for \(I > 0\) almost every orbit initialized in \(\mathcal{O}_1\)
will reach the line \(|q_u| = \pi\).
Since the definition of energy gain only guarantees that orbits initialized in
\(\mathcal{O}_1\) will escape compact subsets of \(\mathcal{O}_1\) in finite
time, we must first prove that almost every orbit in \(\mathcal{O}_1\) will
escape \(\mathcal{O}_1\) through its boundary, the level set \(E_\pi\) with
energy \(E(\pi,0)\).

Define \(\bar{\mathcal{O}}_1 = \mathcal{O}_1 \cup E_\pi\) to be the closure of
\(\mathcal{O}_1\).
An orbit of the acrobot will only begin rotating if it escapes
\(\bar{\mathcal{O}}_1\) by crossing through \(E_\pi\), and eventually remains
outside this set.
Taking the Jacobian of the constrained dynamics
~\eqref{eqn:acrobot-constrained-dynamics} at the upright equilibrium 
\((q_u,p_u) = (\pi,0)\) yields
\[
    J = \begin{bmatrix}
        -\frac{6mgl\bar{q}_aI}{5} & \frac{1 - 2m^2gl^3\bar{q}_a^2 I^2}{5ml^2} \\
        3mgl & mgl\bar{q}_aI
    \end{bmatrix}
    ,
\]
which has characteristic polynomial
\[
    \det\left(\lambda \Id{2} - J\right)
    = \lambda^2 + \frac{mgl\bar{q}_a I}{5} \lambda - 3g
    .
\]
By Descartes' rule of signs, this polynomial always has exactly one root with
positive real part. 
The equilibrium \((\pi,0)\) is therefore unstable, so the stable manifold
\(\Pi^+\) of initial conditions converging to \((\pi,0)\) is one-dimensional,
and hence is of measure zero in \(\SxR\).

Using \(x(t) := (q_u(t),p_u(t))\) as shorthand, let
\(x(0) \in \mathcal{O}_1\) be a nonzero initial condition of the acrobot.
Suppose by way of contradiction that the orbit \(x(\R)\) is confined within 
\(\bar{\mathcal{O}}_1\), and does not exit through \(E_\pi\) into the rotation
zone.
%Since \(\bar{\mathcal{O}}_1\) is compact, the Birkhoff Theorem \cite{birkhoff}
%implies:
%\begin{itemize}
%    \item The positive limit set \(L_+\) of \(x(t)\) is non-empty, compact, and
%        invariant.
%    \item The solution \(x(t)\) asymptotically tends to \(L_+\).
%\end{itemize}
Since the acrobot gains energy on \(\mathcal{O}_1\) and \(\bar{\mathcal{O}}_1\)
is compact, the Birkhoff Theorem \cite{birkhoff} implies that
the solution \(x(t)\) asymptotically tends to the positive limit set, which is
the largest invariant subset of \(E_\pi\).
Taking the limit as \(r \to \pi\) of
\eqref{eqn:theta-dot-nom-osc}--\eqref{eqn:theta-dot-limit},
we get \(\dot{\theta} \geq 0\) on \(E_\pi\) with equality if and only if
\(\theta \in \{0,\pi\}\).
Thus, the largest invariant subset of \(E_\pi\) is either the upright
equilibrium \(\{(\pi,0)\}\) or \(E_\pi\) itself.

Taking the derivative of \(E(q_u,p_u)\) at the \(p_u\)-axis yields
\(\dot{E}(0,p_u) = g/(5l) \sin(q_a)p_u\).
This is non-zero on \(E_\pi\), so \(E_\pi\) is not invariant and the positive
limit set of \(x(t)\) must be the set \(\{(\pi,0)\}\). 
Hence, \(x(t)\) converges to \((\pi,0)\), so \(x(0) \in \Pi^+\).
Since \(\Pi^+\) is a set of measure zero in \(\SxR\), almost every orbit
initialized in \(\mathcal{O}_1\) must escape at least once through \(E_\pi\)
into the rotation domain.

Using this fact, we now prove that almost every orbit in \(\mathcal{O}_1\) will
cross the line \(|q_u| = \pi\).
Let \(x(0) \in \mathcal{O}_1 \backslah \Pi^+\) and suppose that the orbit
\(x([0,\infty[)\) does not hit the line \(|q_u| = \pi\).
This orbit must remain in the compact set \(\Sone \times [-K,K]\) for some 
\(K > 0\).
Indeed, suppose otherwise.
Then for arbitrary \(K > 0\), there exists some \(t_K > 0\) such that
\(|p_u(t_K)| = K\), and there exists a \(t_{K/2}\) so that
\(|p_u(t)| >= K/2\) for \(t \in [t_K, t_K + t_{K/2}]\).
Looking at the constrained dynamics \eqref{eqn:acrobot-constrained-dynamics} in
this timespan, we see that \(\dot{p}_u \geq -3mgl\) and
\(\dot{q}_u \geq c_1 p_u + c_2 \geq c_1 K/2 + c_2\) for some constants
\(c_1, c_2 > 0\).
Picking \(K\) large enough guarantees that \(q_u\) covers a distance of \(2\pi\)
over this time interval, and so the orbit hits the line \(|q_u| = \pi\)
(thus providing a contradiction).

Thus, the orbit \(x(t)\) remains within the compact set \(\Sone \times [-K,K]\).
By the Birkhoff theorem, \(x(t)\) approaches a positive limit set, which is
either a closed orbit or the equilibrium \((\pi,0)\).

TODO: closed orbit case

We conclude that almost all orbits beginning in \(\mathcal{O}_1\) will, in
finite time, escape the closure of \(\mathcal{O}_1\) and begin rotating by
crossing the vertical line \(|q_u| = \pi\).
